---
title: "Cyclistic Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
    number_sections: false
editor_options: 
  markdown: 
    wrap: 72
---

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Imported Data

------------------------------------------------------------------------

To begin my final analysis of the Cyclistic Bike-Share company data, I
imported the data I cleaned and processed in MySQL into R to preform my
analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Loaded in the libraries that I would be using in this analysis
library(dplyr)
library(ggplot2)
library(tidyverse)
library(plotly)
library(leaflet) 
library("viridis")
library(readr)
library(scales)
```

```{r}
cyclistic.cleaned.data <- read.csv("cyclistic.cleaned.data.csv")
locations <- read.csv("locations.csv")
```

```{r}
# Renamed the columns in the data set to match the SQL script
colnames(cyclistic.cleaned.data) <- c("ride_id", "rideable_type", "time_trip_started", "time_trip_ended", "bike_station_start", "start_station_id", "bike_station_end", "end_station_id", "start_lat", "start_lng", "end_lat", "end_lng", "customer_type", "trip_duration_minutes", "weekday", "month", "time_of_day")
cyclistic.cleaned.data$customer_type <- gsub('[\n]','',cyclistic.cleaned.data$customer_type)
```

\

# Summary Statistics

------------------------------------------------------------------------

I began my final analysis by examining the summary statistics for the
whole Cyclistic data set to examine the breakdown in bike usage.

```{r}
(cyclistic.cleaned.data %>%
   summarise(Ride.Count = n(), Average.Trip.Duration = mean(trip_duration_minutes), Classic.Bike.Count = sum(with(cyclistic.cleaned.data, rideable_type== "classic_bike")), Electric.Bike.Count = sum(with(cyclistic.cleaned.data, rideable_type== "electric_bike")), Docked.Bike.Count = sum(with(cyclistic.cleaned.data, rideable_type== "docked_bike"))))
```

\

# Data Visualization

------------------------------------------------------------------------

I then partitioned my data set by customer type; separating annual
members and causal riders, to begin to identify the difference in the
bike-share usage.

## [Member & Casual Analysis]{.underline}

I first began to examine the split in customer type by displaying the
split in customer type usage, and then visualized the difference in
bike trips, time, and bike usage by customer type.

```{r, fig.width = 10,fig.height = 6}
(cyclistic.cleaned.data %>%
  summarise(Casual= sum(with(cyclistic.cleaned.data, customer_type== "casual")), Member = sum(with(cyclistic.cleaned.data, customer_type== "member"))))

# Graphing the split in bike usage and the average length of trips by customer type 
cust <- ggplot(cyclistic.cleaned.data, aes(customer_type, fill = customer_type)) + geom_bar() + labs(x="Customer Type", y="Number of Rides",title="Number of Rides by Customer Type", fill='Customer Type') +  scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "F", end = .48) + scale_x_discrete(labels=c("casual" = "Casual", "member" = "Member"))
(ggplotly(cust))

avg_cust <- ggplot(cyclistic.cleaned.data, aes(x=customer_type, fill = customer_type)) + labs(x="Customer Type", y="Average Trip Duration",title="Average Trip Duration by Customer Type", fill='Customer Type')  +  scale_y_continuous(labels = scales::comma) + stat_summary(aes(y=trip_duration_minutes), fun = "mean", geom = "bar") +  scale_fill_viridis(discrete = TRUE, option = "B", end = .48) + scale_x_discrete(labels=c("casual" = "Casual", "member" = "Member"))
(ggplotly(avg_cust))

# Examining the split in ride type by customer plan, then plotting the observed split

ride_plot <- ggplot(data=cyclistic.cleaned.data, aes(x=rideable_type, fill=customer_type)) + geom_bar(position="dodge", stat = "count") + labs(x="Bike Type", y="Number of Rides",title="Number of Rides for Bike Types by Customer Type", fill='Customer Type') +  scale_y_continuous(labels = scales::comma) +  scale_fill_viridis(discrete = TRUE, option = "B", end = .48) + scale_x_discrete(labels=c("classic_bike" = "Classic Bike", "docked_bike" = "Docked Bike", "electric_bike" = "Electric Bike"))
(ggplotly(ride_plot))

avg_ride <- ggplot(data=cyclistic.cleaned.data, aes(x=rideable_type, fill=customer_type))+ stat_summary(aes(y=trip_duration_minutes), fun = "mean", geom = "bar", position = position_dodge(0.9)) + labs(x="Bike Type", y="Average Trip Duration",title="Average Trip Duration for Bike Types by Customer Type", fill='Customer Type') +  scale_y_continuous(labels = scales::comma) +  scale_fill_viridis(discrete = TRUE, option = "B", end = .48) + scale_x_discrete(labels=c("classic_bike" = "Classic Bike", "docked_bike" = "Docked Bike", "electric_bike" = "Electric Bike"))
(ggplotly(avg_ride))
```

After this I partitioned the data set by time of day, weekday,and month
and visualized this data.

## [Time of Day]{.underline}

To inspect the differences by time of day, I examined the amount of
trips and the average time duration for every hour.

```{r, fig.width = 10,fig.height = 6 }
# Count of Cyclistic's bike trips for each hour of the day.
(time_mode <- cyclistic.cleaned.data %>%
    summarise('12AM' = sum(with(cyclistic.cleaned.data, time_of_day== "12AM")), '1AM' = sum(with(cyclistic.cleaned.data, time_of_day== "1AM")),'2AM' = sum(with(cyclistic.cleaned.data, time_of_day== "2AM")), '3AM' = sum(with(cyclistic.cleaned.data, time_of_day== "3AM")),'4AM' = sum(with(cyclistic.cleaned.data, time_of_day== "4AM")),'5AM' = sum(with(cyclistic.cleaned.data, time_of_day== "5AM")), '6AM' = sum(with(cyclistic.cleaned.data, time_of_day== "6AM")), '7AM' = sum(with(cyclistic.cleaned.data, time_of_day== "7AM")), '8AM' = sum(with(cyclistic.cleaned.data, time_of_day== "8AM")), '9AM' = sum(with(cyclistic.cleaned.data, time_of_day== "9AM")),'10AM' = sum(with(cyclistic.cleaned.data, time_of_day== "10AM")),'11AM' = sum(with(cyclistic.cleaned.data, time_of_day== "11AM")), '12PM' = sum(with(cyclistic.cleaned.data, time_of_day== "12PM")), '1PM' = sum(with(cyclistic.cleaned.data, time_of_day== "1PM")),'2PM' = sum(with(cyclistic.cleaned.data, time_of_day== "2AM")), '3PM' = sum(with(cyclistic.cleaned.data, time_of_day== "3AM")),'4PM' = sum(with(cyclistic.cleaned.data, time_of_day== "4AM")),'5PM' = sum(with(cyclistic.cleaned.data, time_of_day== "5AM")),'6PM' = sum(with(cyclistic.cleaned.data, time_of_day== "6AM")),'7PM' = sum(with(cyclistic.cleaned.data, time_of_day== "7AM")),'8PM' = sum(with(cyclistic.cleaned.data, time_of_day== "8AM")),'9PM' = sum(with(cyclistic.cleaned.data, time_of_day== "9AM")),'10PM' = sum(with(cyclistic.cleaned.data, time_of_day== "10AM")),'11PM' = sum(with(cyclistic.cleaned.data, time_of_day== "11PM"))))

name_time <- c('12AM', '1AM', '2AM', '3AM', '4AM', '5AM', '6AM', '7AM', '8AM', '9AM', '10AM', '11AM', '12PM', '1PM', '2PM', '3PM', '4PM', '5PM', '6PM', '7PM', '8PM', '9PM', '10PM', '11PM')
cyclistic.cleaned.data$time_of_day <- factor(cyclistic.cleaned.data$time_of_day, levels = name_time)

# Graphing of the count of bike trips for each hour of the day and the average bike trip duration for each hour of the day

time <- ggplot(cyclistic.cleaned.data, aes(x = time_of_day, fill = time_of_day)) + geom_bar() + labs(x="Time of Day", y="Number of Rides",title = "Number of Rides by the Time of Day (AM/PM)", fill = 'Time of Day (AM/PM)') + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B") + theme(axis.text.x=element_text(angle=30, vjust=1, hjust=1))
(ggplotly(time))

avg_time <- ggplot(data=cyclistic.cleaned.data, aes(x=time_of_day, fill=time_of_day)) + labs(x="Time of Day (AM/PM)", y="Average Trip Duration\nin Minutes",title="Average Trip Duration by the Time of Day (AM/PM)", fill = 'Time of Day (AM/PM)') + stat_summary(aes(y =trip_duration_minutes), fun = "mean", geom = "bar", position = position_dodge(width = .9)) + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B") + theme(axis.text.x=element_text(angle=30, vjust=.8, hjust=0.8))
(ggplotly(avg_time))
```

I then look into the difference in customer type by examining the amount
of trips and the average time duration for every hour by customer type.

```{r, fig.width = 10,fig.height = 6 }
cust_time <- ggplot(cyclistic.cleaned.data, aes(x=time_of_day, fill=customer_type)) + geom_bar(position="dodge", stat = "count") + labs(x="Time of Day", y="Number of Rides", title="Number of Rides by Time of Day (AM/PM)\nby Customer Type", fill='Customer Type') + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "F", end = .48) + theme(axis.text.x=element_text(angle=30, vjust=.8, hjust=0.8))
(ggplotly(cust_time))

avg_cust_time <- ggplot(data=cyclistic.cleaned.data, aes(x=time_of_day, fill=customer_type)) + labs(x="Time of Day", y="Average Trip Duration\nin Minutes",title="Average Trip Duration by Time of Day (AM/PM)\nby Customer Type", fill='Customer Type') + scale_y_continuous(labels = scales::comma) + stat_summary(aes(y =trip_duration_minutes ), fun = "mean", geom = "bar", position = position_dodge(width = .9)) + scale_fill_viridis(discrete = TRUE, option = "F", end = .48) + theme(axis.text.x=element_text(angle=30, vjust=.8, hjust=0.8))
(ggplotly(avg_cust_time))
```

## [Week]{.underline}

Next I inspected the differences in rider usage by examining the amount
of trips and the average time duration for every weekday.

```{r, fig.width = 10, fig.height = 6}
# Count of Cyclistic's  bike trips for each day of the week.
(weekday_mode <- cyclistic.cleaned.data %>%
    summarise(Monday = sum(with(cyclistic.cleaned.data, weekday== "Monday")), Tuesday = sum(with(cyclistic.cleaned.data, weekday== "Tuesday")), Wednesday = sum(with(cyclistic.cleaned.data, weekday== "Wednesday")), Thursday = sum(with(cyclistic.cleaned.data, weekday== "Thursday")), Friday =  sum(with(cyclistic.cleaned.data, weekday== "Friday")), Saturday = sum(with(cyclistic.cleaned.data, weekday== "Saturday")), Sunday = sum(with(cyclistic.cleaned.data, weekday== "Sunday"))))

name_week <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday')
cyclistic.cleaned.data$weekday <- factor(cyclistic.cleaned.data$weekday, levels = name_week)

# Graphing of the count of bike trips for each day of the week and the average bike trip duration for each day of the week

week <- ggplot(data=cyclistic.cleaned.data, aes(x=weekday, fill=weekday)) + geom_bar(position="dodge", stat = "count") + labs(x="Weekday", y="Number of Rides",title="Number of Rides by the Day of Week", fill = 'Weekday') + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B")
(ggplotly(week))

avg_week <- ggplot(data=cyclistic.cleaned.data, aes(x=weekday, fill=weekday)) + labs(x="Weekday", y="Average Trip Duration\nin Minutes",title="Average Trip Duration by the Day of Week", fill = 'Weekday') + stat_summary(aes(y =trip_duration_minutes), fun = "mean", geom = "bar", position = position_dodge(width = .9)) + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B") 
(ggplotly(avg_week))
```

I again looked into the differences in customer type by examining the
amount of trips and the average trip duration for each weekday by
customer type.

```{r, fig.width = 10, fig.height = 6}
cust_week <- ggplot(data=cyclistic.cleaned.data, aes(x=weekday, fill=customer_type)) + geom_bar(position="dodge", stat = "count") + labs(x="Weekday", y="Number of Rides",title="Number of Rides by Weekday\nby Customer Type", fill='Customer Type') + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B", end = .48) 
ggplotly(cust_week)

avg_cust_week <- ggplot(data=cyclistic.cleaned.data, aes(x=weekday, fill=customer_type)) + labs(x="Weekday", y="Average Trip Duration", title="Average Trip Duration by Weekday\nby Customer Type", fill='Customer Type') + stat_summary(aes(y =trip_duration_minutes ), fun = "mean", geom = "bar", position = position_dodge(width = .9)) + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B", end = .48) 
ggplotly(avg_cust_week)
```

## [Month]{.underline}

Lastly I inspected the differences in rider usage by month by examining
the amount of trips and the average time duration for each month.

```{r, fig.width = 10,fig.height = 6}
# Count of Cyclistic's bike trips for each month
(month_mode <- cyclistic.cleaned.data %>%
summarise(January = sum(with(cyclistic.cleaned.data, month== "January")), February = sum(with(cyclistic.cleaned.data, month== "February")), March = sum(with(cyclistic.cleaned.data, month== "March")), April = sum(with(cyclistic.cleaned.data, month== "April")), May =  sum(with(cyclistic.cleaned.data, month== "May")), June = sum(with(cyclistic.cleaned.data, month== "June")), July = sum(with(cyclistic.cleaned.data, month== "July")), August = sum(with(cyclistic.cleaned.data, month== "August")), September = sum(with(cyclistic.cleaned.data, month== "September")), October = sum(with(cyclistic.cleaned.data, month== "October")), November = sum(with(cyclistic.cleaned.data, month== "November")), December = sum(with(cyclistic.cleaned.data, month== "December"))))

name_month <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")
cyclistic.cleaned.data$month <- factor(cyclistic.cleaned.data$month, levels = name_month)

# Graphing of the count of bike trips for each Month and the bike trip duration for each Month.

month <- ggplot(cyclistic.cleaned.data, aes(month, fill=month)) + geom_bar() + labs(x="Month", y="Number of Rides",title="Number of Rides by the Month", fill='Month') + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B") + theme(axis.text.x=element_text(angle=30, vjust=.8, hjust=0.8))
(ggplotly(month))

avg_month <- ggplot(data=cyclistic.cleaned.data, aes(x=month, fill=month)) + stat_summary(aes(y =trip_duration_minutes), fun = "mean", geom = "bar", position = position_dodge(width = .9)) + labs(x="Month", y="Average Trip Duration\nin Minutes",title="Average Trip Duration by the Month", fill='Month') + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B") + theme(axis.text.x=element_text(angle=30, vjust=.8, hjust=0.8))
(ggplotly(avg_month))
```

I then looked into the differences in customer type by month examining
the amount of trips and the average time duration for each month by
customer type.

```{r, fig.width = 10,fig.height = 6}
cust_month <- ggplot(data=cyclistic.cleaned.data, aes(x=month, fill=customer_type)) + geom_bar(position="dodge", stat = "count") + labs(x="Month", y="Number of Rides",title="Number of Rides by Month\nby Customer Type", fill='Customer Type') + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "F", end = .48) + theme(axis.text.x=element_text(angle=30, vjust=.8, hjust=0.8))
(ggplotly(cust_month))

avg_cust_month <- ggplot(data=cyclistic.cleaned.data, aes(x=month, fill=customer_type)) + labs(x="Month", y="Average Number of Rides",title="Average Trip Duration by Month\nby Customer Type", fill='Customer Type') + stat_summary(aes(y =trip_duration_minutes ), fun = "mean", geom = "bar", position = position_dodge(width = .9)) + scale_y_continuous(labels = scales::comma) + scale_fill_viridis(discrete = TRUE, option = "B", end = .48) + theme(axis.text.x=element_text(angle=30, vjust=.8, hjust=0.8))
(ggplotly(avg_cust_month))
```

\

# Popular/Unpopular Bike Stations

------------------------------------------------------------------------

Finally I examined Cyclistic's bike station locations, and found the
most popular and least popular bike stations; then plotted the location
of these stations onto a map of the city of Chicago.

```{r, out.width="100%", warning = FALSE}
# Renamed the columns in the data set to match the SQL script
colnames(locations) <- c("bike_station_combine", "count", "combine_lng", "combine_lat")

mybins <- seq(0, 80000, by = 20000)

#Creating hover text for map 
mytext <- paste(
   "Bike Station Name: ", locations$bike_station_combine, "<br/>", 
   "Count of Trips: ", locations$count, sep="") %>%
  lapply(htmltools::HTML)

#Creating of color palette for location data on map
mypalette<- colorBin(palette="inferno", domain=locations$count, na.color="transparent", bins=mybins, reverse = TRUE)

#Creation of map using leaflet package
map <- leaflet(locations) %>% 
  addTiles() %>%  
  
# Centering the map over the city of Chicago
  setView(
    lng = -87.66984677949904, lat = 41.89617798482842,  zoom = 11.5
        ) %>%
  addProviderTiles("CartoDB.Positron") %>%
  
# Plotting each bike station location on the map, and counting the usage of the the bike station
  addCircleMarkers(
    ~ combine_lng, ~ combine_lat, 
    fillColor = ~ mypalette(count), 
    fillOpacity = 0.7, 
    color = "white", 
    radius = 8, 
    stroke = FALSE,
# Adding in text to the map to identify the name of the bike station and its usage
    label = mytext,
    labelOptions = labelOptions(
      style = list( 
        "font-weight" = "normal", 
        padding = "3px 8px"
        ), 
      textsize = "13px", 
      direction = "auto"
      ) 
    ) %>%
  addLegend( 
    pal = mypalette, 
    values = locations$count, 
    opacity = 0.9,
    title = "Count of Trips", 
    position = "bottomright"
    )
map
```

## [Top Bike Stations]{.underline}

Identifying and mapping the top ten percent of Cyclistic's bike station locations

```{r, out.width="100%", warning = FALSE}
# Created a new dataset containing Cyclistic's fifty most used bike station locations. 

(locations_top <- locations %>%                
  arrange(desc(count)) %>% 
  group_by(bike_station_combine) %>%
    rename("Bike.Station" = "bike_station_combine", "Count" = "count", "Longitude" = "combine_lng", "Latitude" = "combine_lat") %>%
    head(120))

mybins1 <- seq(30000, 80000, by = 10000)
mypalette1<- colorBin( palette="inferno", domain=locations_top$Count, na.color="transparent", bins=mybins1, reverse = TRUE)

mytext1 <- paste(
   "Bike Station Name: ", locations_top$Bike.Station, "<br/>", 
   "Count of Trips: ", locations_top$Count, sep="") %>%
  lapply(htmltools::HTML)

(map <- leaflet(locations_top, width = "100%") %>% 
  addTiles() %>%  
  setView(
    lng = -87.63875, lat = 41.90419,  zoom = 11.5
        ) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
   ~ Longitude, ~ Latitude,  
    fillColor = ~ mypalette1(Count), 
    fillOpacity = 0.7, 
    color = "white", 
    radius = 8, 
    stroke = FALSE,
    label = mytext1,
    labelOptions = labelOptions(
      style = list( 
        "font-weight" = "normal", 
        padding = "3px 8px"
        ), 
      textsize = "13px", 
      direction = "auto"
      ) 
    ) %>%
  addLegend( 
    pal = mypalette1, 
    values = locations_top$Count, 
    opacity = 0.9,
    title = "Count of Trips", 
    position = "bottomright"
    ))
```

## [Bottom Bike Stations]{.underline}

Identifying and mapping the bottom ten percent of Cyclistic's bike station locations

```{r, out.width="100%", warning = FALSE}
# Created a new dataset containing Cyclistic's fifty least used bike station locations. 

(locations_bot <- locations %>%                
  arrange(count) %>% 
  group_by(bike_station_combine) %>%
  rename("Bike.Station" = "bike_station_combine", "Count" = "count", "Longitude" = "combine_lng", "Latitude" = "combine_lat") %>%
  head(120))

mybins2 <- seq(0, 5, by = 1)
mypalette2<- colorBin(palette="inferno", domain=locations_bot$Count, na.color="transparent", bins=mybins2, reverse = FALSE)

mytext2 <- paste(
   "Bike Station Name: ", locations_bot$Bike.Station, "<br/>", 
   "Count of Trips: ", locations_bot$Count, sep="") %>%
  lapply(htmltools::HTML)

(map <- leaflet(locations_bot) %>% 
  addTiles() %>%  
  setView(
    lng = -87.75331, lat = 41.84328,  zoom = 11
        ) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
   ~ Longitude, ~ Latitude,  
    fillColor = ~ mypalette2(Count), 
    fillOpacity = 0.7, 
    color = "white", 
    radius = 8, 
    stroke = FALSE,
    label = mytext2,
    labelOptions = labelOptions(
      style = list( 
        "font-weight" = "normal", 
        padding = "3px 8px"
        ), 
      textsize = "13px", 
      direction = "auto"
      ) 
    ) %>%
  addLegend( 
    pal = mypalette2, 
    values = locations_bot$Count, 
    opacity = 0.9,
    title = "Count of Trips", 
    position = "bottomright"
    ))
```
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>