---
title: "Final Report"
output:
  html_document:
    toc: TRUE
    toc_depth: 1
    number_sections: false
    toc_float: true
    df_print: paged
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Objective

------------------------------------------------------------------------

Until now, Cyclistic's marketing strategy relied on building general
awareness and appealing to broad consumer segments. One approach that
helped make these things possible was the flexibility of its pricing
plans; being single-ride passes, full-day passes, and annual
memberships. Customers who purchase single-ride or full-day passes are
referred to as casual riders. Customers who purchase annual memberships
are Cyclistic annual members.

Recently, Cyclistic's finance analysts have concluded that annual
members are much more profitable than casual riders. Although the
pricing flexibility helps Cyclistic attract more customers, Cyclistic's
Director of Marketing Lily Moreno believes that maximizing the number of
annual members will be key to future growth. Therefore rather than
creating a marketing campaign that targets all-new customers, Director
Moreno has set a clear goal:

-   Design marketing strategies aimed at converting casual riders into
    annual members.

In order to do that, the marketing analyst team needs to better
understand how annual members and casual riders differ, why casual
riders would buy a membership, and how digital media could affect their
marketing tactics. Therefore this final report delivers analyst and
recommendations on how Cyclistic can convert existing causal riders into
annual memberships.

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Loaded in the libraries that I would be using in this analysis
library(dplyr)
library(ggplot2)
library(tidyverse)
library(plotly)
library(leaflet) 
library("viridis")
library(readr)

# Loading in data
cyclistic.cleaned.data <- read.csv("cyclistic.cleaned.data.csv")
locations <- read.csv("locations.csv")

colnames(cyclistic.cleaned.data) <- c("ride_id", "rideable_type", "time_trip_started", "time_trip_ended", "bike_station_start", "start_station_id", "bike_station_end", "end_station_id", "start_lat", "start_lng", "end_lat", "end_lng", "customer_type", "trip_duration_minutes", "weekday", "month", "time_of_day")
cyclistic.cleaned.data$customer_type <- gsub('[\n]','',cyclistic.cleaned.data$customer_type)

colnames(locations) <- c("bike_station_combine", "count", "combine_lng", "combine_lat")
```

\

# Analysis & Findings

------------------------------------------------------------------------

## [Bike Ride Analysis]{.underline}

After final analysis of the 5,163,998 bike rides Cyclistic's customers
took over the year-long timespan between July 2021 and July 2022 in R
Studio, I discovered many important findings about Cyclistic's customers
usage of the bike-share program, and the differences between how annual
members and casual riders utilized the bike-share program. To begin I
found that the average customer prefers to ride on the Classic bike, and
the average customer's trip on their bike rental takes approximately 17
minutes to complete.

```{r}
(cyclistic.cleaned.data %>%
   summarise(Average.Trip.Duration = mean(trip_duration_minutes), Classic.Bike.Count = sum(with(cyclistic.cleaned.data, rideable_type== "classic_bike")), Electric.Bike.Count = sum(with(cyclistic.cleaned.data, rideable_type== "electric_bike")), Docked.Bike.Count = sum(with(cyclistic.cleaned.data, rideable_type== "docked_bike"))))
```

Additionally, breaking down further by membership type, we see that
while 2,969,713 riders are annual members, 2,194,285 riders are casual
riders meaning their is huge market of potential new annual member
converts. Moreover looking at the differences in how the different
riders use the bike-share program, we see that while all riders tend to
prefer to use the Classic bike, annual members took much shorter bike
trips at approximately 12 minutes, while casual members bike trip were
much long at approximately 24 minutes.

```{r}
(cyclistic.cleaned.data %>%
   group_by(customer_type) %>%
   arrange(customer_type) %>%
    summarise(Average.Trip.Duration = mean(trip_duration_minutes), Classic.Bike.Count = sum(rideable_type ==  "classic_bike"), Electric.Bike.Count = sum(rideable_type == "electric_bike"), Docked.Bike.Count = sum(rideable_type== "docked_bike")))
```

## [Time Analysis]{.underline}

Furthermore analyzing the time that these trips took place shows that
overall, Cyclistic's customers prefer to ride during the summer months
with July being the most popular month averaging 795,447 rides with bike
usage dropping heavily during the winter. We also see that trip duration
follows this trend going up to \~20 minutes during summer and falling
down to \~13 minutes during the winter. Looking at which weekday
Cyclistic's customers prefer shows that generally customers use the
bike-share program equally throughout the week with a slight preference
for Saturday averaging 861,783 rides. Furthermore, riders also took
slightly longer trips during the weekends rising from around \~15
minutes to \~20 minutes during the end of the week. Lastly checking
overall customer preference along the time of day shows that bike usage
follows a normal distribution as starting around 5AM, an average of 41,816
customers begin use the ride-share program before peaking at around
514,116 customers around 5PM. However despite this trips throughout the
day stayed around \~16 minutes on average only falling for a short
period of time during 5AM-8AM.

```{r}
(cyclistic.cleaned.data %>%
   group_by(month) %>%
   summarise(Count = n(),
  Average.Trip.Duration = mean(trip_duration_minutes))  %>%
   arrange(desc(Count)))
 
(cyclistic.cleaned.data %>%
   group_by(Weekday = weekday) %>%
   summarise(Count = n(),
  Average.Trip.Duration = mean(trip_duration_minutes)) %>% arrange(desc(Count)))

(cyclistic.cleaned.data %>%
   group_by(Time.of.Day = time_of_day) %>%
   summarise(Count = n(),
  Average.Trip.Duration = mean(trip_duration_minutes)) %>% arrange(desc(Count)))

```

Moving back to the breakdown between annual customers and casual riders,
we see that throughout the year annual members take more trips than
casual riders, except during the summer months of July and August. We
see this trend repeated when we look closer, and further separate the
months into weekdays; as annual members generally take more trips during
the week, except during the weekend days of Saturday and Sunday where
casual riders outpace annual members. Moreover as shown across the month
and weekday analysis, causal riders take approximately twice as long
during their bike trips, as across both time analyses, annual members
trip's duration were \~60% shorter than casual riders duration. Lastly
turning towards the analysis of the time of day we get a better
understanding of how both customer types utilize the bike-share program.
Specifically we see that while annual members exceed causal rider
throughout most of the day, and they both follow the same normal
distribution peaking at around 5PM. Lastly, at the end of the day
starting at 10PM casual riders overtake annual members till early
morning around 5AM.

```{r}
(cyclistic.cleaned.data %>%
   group_by(Customer.Type = customer_type, Month = month) %>%
   summarise(Count = n(),
  Average.Trip.Duration = mean(trip_duration_minutes))  %>%
   arrange(desc(Count), .by_group=TRUE))
 
(cyclistic.cleaned.data %>%
   group_by(Customer.Type = customer_type, Weekday = weekday) %>%
   summarise(Count = n(),
  Average.Trip.Duration = mean(trip_duration_minutes)) %>% 
    arrange(desc(Count), .by_group=TRUE))

(cyclistic.cleaned.data %>%
   group_by(Customer.Type = customer_type, Time.of.Day = time_of_day) %>%
   summarise(Count = n(),
  Average.Trip.Duration = mean(trip_duration_minutes)) %>% 
    arrange(desc(Count), .by_group=TRUE))

```

## [Geo-spatial Analysis]{.underline}

Finally, analyzing all of Cyclistic's bike station in Chicago indicated
a huge variance in usage between each of the bike station locations;
indicating areas in Chicago where Cyclistic's marketing team can focus
their marketing efforts to maximize customer engagement, and areas for
Cyclistic's marketing team to avoid. To begin, analyzing the map of all
the Cyclistic's bike stations across Chicago displays that customers
preferred bike stations along the eastern coast of Chicago with the bike
stations along Chicago's periphery being more unpopular with Cyclistic's
customers.

```{r, out.width="100%"}
mybins <- seq(0, 80000, by = 20000)
#Creating hover text for map 
mytext <- paste(
   "Bike Station Name: ", locations$bike_station_combine, "<br/>", 
   "Count of Trips: ", locations$count, sep="") %>%
  lapply(htmltools::HTML)
#Creating of color palette for location data on map
mypalette<- colorBin(palette="inferno", domain=locations$count, na.color="transparent", bins=mybins, reverse = TRUE)
#Creation of map using leaflet package
(map <- leaflet(locations) %>% 
  addTiles() %>%  
# Centering the map over the city of Chicago
  setView(
    lng = -87.66984677949904, lat = 41.89617798482842,  zoom = 11.5
        ) %>%
  addProviderTiles("CartoDB.Positron") %>%
# Plotting each bike station location on the map, and counting the usage of the the bike station
  addCircleMarkers(
    ~ combine_lng, ~ combine_lat, 
    fillColor = ~ mypalette(count), 
    fillOpacity = 0.7, 
    color = "white", 
    radius = 8, 
    stroke = FALSE,
# Adding in text to the map to identify the name of the bike station and its usage
    label = mytext,
    labelOptions = labelOptions(
      style = list( 
        "font-weight" = "normal", 
        padding = "3px 8px"
        ), 
      textsize = "13px", 
      direction = "auto"
      ) 
    ) %>%
  addLegend( 
    pal = mypalette, 
    values = locations$count, 
    opacity = 0.9,
    title = "Count of Trips", 
    position = "bottomright"
    ))
```

Diving further into this analysis demonstrates that while the previous
map shows that most of Cyclistic's customers use the bike-share program
within the dense area of bike stations close to the city center;
analyzing the top 10% of Cyclistic's bike stations shows that
Cyclistic's most used bike stations are on the north side of Chicago's
city center that aligns with the area that contains the major tourist
attractions in Chicago. Furthermore the popularity of the bike stations
with close proximity to Chicago's coast suggests many customers use
Cyclistic for scenic bike riding along the coast. Therefore, in order to
maximize the number of annual membership; Cyclistic's marketing analyst
team should focus any campaign strategy or efforts in Chicago's city
center, and the eastern and northern regions bordering Chicago's city
center.

```{r, out.width="100%", warning = FALSE}
(locations_top <- locations %>%                
  arrange(desc(count)) %>% 
  group_by(bike_station_combine) %>%
    rename("Bike.Station" = "bike_station_combine", "Count" = "count", "Longitude" = "combine_lng", "Latitude" = "combine_lat") %>%
    head(120))
mybins1 <- seq(30000, 80000, by = 10000)
mypalette1<- colorBin( palette="inferno", domain=locations_top$Count, na.color="transparent", bins=mybins1, reverse = TRUE)

mytext1 <- paste(
   "Bike Station Name: ", locations_top$Bike.Station, "<br/>", 
   "Count of Trips: ", locations_top$Count, sep="") %>%
  lapply(htmltools::HTML)

(map <- leaflet(locations_top, width = "100%") %>% 
  addTiles() %>%  
  setView(
    lng = -87.63875, lat = 41.90419,  zoom = 11.5
        ) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
   ~ Longitude, ~ Latitude,  
    fillColor = ~ mypalette1(Count), 
    fillOpacity = 0.7, 
    color = "white", 
    radius = 8, 
    stroke = FALSE,
    label = mytext1,
    labelOptions = labelOptions(
      style = list( 
        "font-weight" = "normal", 
        padding = "3px 8px"
        ), 
      textsize = "13px", 
      direction = "auto"
      ) 
    ) %>%
  addLegend( 
    pal = mypalette1, 
    values = locations_top$Count, 
    opacity = 0.9,
    title = "Count of Trips", 
    position = "bottomright"
    ))

```

Lastly, analyzing solely the bottom 10% of Cyclistic's bike stations
shows that customers tended not to used the Cyclistic's bike stations
that where away from the city center of Chicago with specifically the areas to the south and
northwest of Chicago's city center being used the least. Therefore,
Cyclistic's marketing analyst team should avoid focusing their marketing
efforts in these areas.

```{r, out.width="100%", warning = FALSE}
(locations_bot <- locations %>%                
  arrange(count) %>% 
  group_by(bike_station_combine) %>%
  rename("Bike.Station" = "bike_station_combine", "Count" = "count", "Longitude" = "combine_lng", "Latitude" = "combine_lat") %>%
  head(120))

mybins2 <- seq(0, 5, by = 1)
mypalette2<- colorBin(palette="inferno", domain=locations_bot$Count, na.color="transparent", bins=mybins2, reverse = FALSE)

mytext2 <- paste(
   "Bike Station Name: ", locations_bot$Bike.Station, "<br/>", 
   "Count of Trips: ", locations_bot$Count, sep="") %>%
  lapply(htmltools::HTML)

(map <- leaflet(locations_bot) %>% 
  addTiles() %>%  
  setView(
    lng = -87.75331, lat = 41.84328,  zoom = 11
        ) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
   ~ Longitude, ~ Latitude,  
    fillColor = ~ mypalette2(Count), 
    fillOpacity = 0.7, 
    color = "white", 
    radius = 8, 
    stroke = FALSE,
    label = mytext2,
    labelOptions = labelOptions(
      style = list( 
        "font-weight" = "normal", 
        padding = "3px 8px"
        ), 
      textsize = "13px", 
      direction = "auto"
      ) 
    ) %>%
  addLegend( 
    pal = mypalette2, 
    values = locations_bot$Count, 
    opacity = 0.9,
    title = "Count of Trips", 
    position = "bottomright"))
```

\

# Recommendations

------------------------------------------------------------------------

Finally, as stated earlier the purpose of this analysis was to identify
trends to better understand how annual members and casual riders differ
to better inform marketing strategies aimed at converting casual riders
into annual members; and as a result of my analysis I have generated
three recommendations for the Cyclistic's marketing analyst team that
outline how to convert Cyclistic's existing causal riders into annual
memberships.

The three marketing recommendations for the marketing analyst team are
outlined below:

1.  ***A Digital Ad Campaign highlighting the Tourist Attractions in
    Chicago's Metro Hub and Coast***

The final analysis of Cyclistic's bike stations locations displayed that
riders' most popular bike stations were stationed near Chicago's metro
hub and coast, validating prior knowledge that most riders use the
bike-share program for leisure. Moreover deeper analysis into causal
riders supported this analysis as well, as analysis showed that causal
riders preferred to ride on the weekend and after regular work hours
suggesting casual riders were riding for fun rather than for commuting
to work. Therefore, the Cyclistic marketing team should capitalized on
this by running an digital ad campaign targeted towards causal riders
that both highlights the major tourist attractions in metro Chicago and
scenic coastline, and promotes the use of Cyclistic's bikes. Finally this
could be accomplished by combining promotional ads that display
Cyclistic's customers enjoying their time on bikes at major metro
Chicago spots, and by creating a series of suggested tailored scenic
bike routes through metro Chicago that encourage riders to try the
different bike routes over a extended period of time. Thus, encouraging
casual riders to switch to the annual membership in order to save on the
multiple scenic bike rides.

2.  ***A Digital Marketing Campaign highlighting a Discount on Electric
    Bikes for Annual Members***

Furthermore, while most riders use the bike-share program for leisure,
analysis showed that causal riders overtake annual members during the
late night hours of 10PM till 5AM, suggesting that these later riders
were using the bike for commuting and travel. Additionally, across all
time casual riders take bike trips that are significantly longer than
annual members, therefore I recommend a targeted marketing campaign
promoting a discount on electric bikes for annual members, that would
run digital ads at bike stations at night. Thus, casual riders would be
incentivize to convert to annual memberships to attain the benefits of
shorter trips times and no longer needing to cycling from using electric bikes,
which would be especially beneficial to this section of
casual riders traveling or commuting at night after a long tiring day.

3.  ***Offer Discounted Cyclistic Annual Memberships during the
    Summer***

Finally the Cyclistic analyst marketing team should take advantage of
the significant increase in causal riders during the summer season by
offering an promotional discount that lowers the price of the annual
membership during the month of July, August, and September. Furthermore
as a result of Chicago's clement weather during this time, Cyclistic can
also capitalize on the increase in ridership and good weather by
partnering the discount with ads highlighting serene bike rides in the
pleasant weather.
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>